Bayesian Sparsification of Deep $\cplx$-valued networks

### 2020-02-05 paper/paper.tex
minor on-char fixes

With a special family of posterior approximation, variational inference can be used as
a regularization method and as a model sparsification technique. \citet{kingma_variational_2015}
consider Dropout \citep{hinton_improving_2012}, DropConnect \citep{wan_regularization_2013},
and Gaussian dropout \citep{srivastava_dropout_2014,wang_fast_2013} through the lens
of Bayesian inference methods and proposes \textit{Variational Dropout}. They argue that
the multiplicative noise introduced by these methods induces a distribution equivalent to
a fully factorized variational posterior of the form $
  q_\theta(\omega) = \prod_j q_{\theta}(\omega_j)
$, where $q_{\theta}(\omega_j)$ is $\omega_j = \mu_j \xi_j$ with $
  \xi_j \sim p_\theta(\xi_j)
$ iid from some $p_\theta(\xi)$.
%
% Binary dropout with rate $p \in (0, 1)$ uses $
%   p_\xi
%     = \mathcal{B}\bigl(
%       \{0, \tfrac1{1-p}\}, 1-p
%     \bigr)
% $, whereas in Gaussian dropout $
%   \xi \sim \mathcal{N}(1, \alpha)
% $ for $\alpha = \tfrac{p}{1-p}$.
% This suggests making $\alpha$ a variational parameter
% and optimizing it in \eqref{eq:elbo_general} with an appropriate penalty term. Thus,
%
Variational Dropout assumes fully factorized Gaussian approximate posterior $
  q_\theta(\omega)
    = \prod_j \mathcal{N}(\mu_j, \alpha_j \mu_j^2)
$ and factorized log-uniform prior $\pi(\omega)$ with $
  \pi(\omega_j) \propto \lvert \omega_j \rvert^{-1}
$. The divergence term in \eqref{eq:elbo} unravels into a sum of the following
analytically intractable terms:
\begin{equation}  \label{eq:improper-kl-div-real}
  KL(q_\theta \| \pi)
    \propto
      \frac12 \mathbb{E}_{\varepsilon \sim \mathcal{N}(0, 1)}
        \log{\Bigl\lvert \tfrac1{\sqrt{\alpha}} + \varepsilon \Bigr\rvert^2}
      % \frac12 \mathbb{E}_{\varepsilon \sim \mathcal{N}(0, 1)}
      %   \log{\bigl\lvert 1 + \sqrt{\alpha} \varepsilon \bigr\rvert^2}
      % - \frac12 \log{\alpha}
  \,.
\end{equation}
\citet{kingma_variational_2015} approximate this expression by a non-linear polynomial
regression for truncated $\alpha \in (0, 1)$. \citet{molchanov_variational_2017}
refine the approximation by using a nonlinear regression based on sigmoid and soft-plus.
% see p.6 sec 3.4 of kingma_variational_2015
We independently verify their approximation against Monte-Carlo estimates for $\alpha$
varying over a fine logarithmic grid and the exact expression for the derivative of
\eqref{eq:improper-kl-div-real}, which we derive in appendix~\ref{sub:real-chisq-grad}.
% (!) good for SGVB: gradient-based methods require unbiased gradient estimates

@@

\begin{figure}[ht]
  \centering
  \includegraphics[draft,width=1.\linewidth]{../assets/log-alpha__sum__fashion-mnist.pdf}
  \caption{%
    Smoothed density of $\log\alpha$ for a convolutional and dense layer from the
    model in sec.~\ref{sub:mnist_like_datasets}.
  }
  \label{fig:log-alpha__mnist-like}
\end{figure}

% on compression threshold
The threshold $\tau$ is picked so, that the remaining \textit{``non-zero''} parameters
are within $\delta$ relative tolerance of their mean with high probability under the
approximate posterior $q_\theta(W)$. Since \eqref{eq:c-gauss-vi} is Gaussian, $
  \tfrac{\lvert w - \mu \rvert^2}
        {\alpha \lvert \mu \rvert^2}
$ is $\chi^2_k$ distributed with shape $k=1$ ($\real$) or $2$ ($\cplx$), whence
\begin{equation} \label{eq:prob_relevant}
  \Pr\bigl(
    \lvert w - \mu \rvert \leq \delta \lvert \mu \rvert
  \bigr)
    = \Pr\biggl(
      \frac{\lvert w - \mu \rvert^2}{\alpha \lvert \mu \rvert^2}
        \leq \frac{\delta^2}{\alpha}
    \biggr)
    = \chi^2_k\biggl(\frac{\delta^2}{\alpha}\biggr)
  \,.
\end{equation}
% Figure~\ref{fig:log-alpha__mnist-like} (bottom) shows the probability of a $\real$
% (or $\cplx$)  parameter being non-zero for various $\delta\in(0, \tfrac12)$ as a function
% of $\log \alpha$.
For a reasonably generous $\delta = 25\%$ tolerance all values of $\log \alpha$ below
$-2.5$ yield a relatively high chance of a non-zero parameter. Given that $q_\theta$
is merely an approximation, we opt for a somewhat higher $\log \alpha$ threshold, and
pick $\tau = -\tfrac12$, which is much lower than $\tau = 3$, recommended by
\citet{molchanov_variational_2017}.

@@

% mention the same experimental setup as in MNIST
Experiments on CIFAR10 dataset are set up similarly to MNIST-like experiment, except
that we study VGG16 model \citep{simonyan_very_2015} and its $\cplx$-counterpart, fit
on the full training split of $60k$ colour $32\times32$ images, and use only raw image
input features (\texttt{raw}).

@@

\begin{figure}[!ht]
  \centering
  \begin{subfigure}[b]{0.9\textwidth}  % imcl2019-style swears at this
    \centering
    \includegraphics[draft,width=1.\linewidth]{../assets/grid_cplx_fine_kl_div_v3_fast-c-ard-trade-off.pdf}
    % \caption{} \label{fig:log-alpha-kde}
  \end{subfigure} \\
  % \hfill
  \begin{subfigure}[b]{0.9\textwidth}  % imcl2019-style swears at this
    \centering
    \includegraphics[draft,width=1.\linewidth]{../assets/grid_cplx_fine_kl_div_v3_fast-c-ard-trade-off.pdf}
    % \caption{} \label{fig:log-alpha-proba}
  \end{subfigure}
  \caption{%
    Pre ``finte-tune'' performance of the best models from figure \ref{fig:musicnet_trade-off}
    compressed with methods of sec.~\ref{sec:dropout_for_c_valued_layers}:
    (\ref{ssub:variational_dropout} -- \textit{top}, \ref{ssub:empirical_bayes} --  \textit{bottom}).
  }
  \label{fig:musicnet_trade-off_pre_fune-tune}
\end{figure}
% subsection muscinet (end)

>>>>

With a special family of posterior approximation, variational inference can be used as
a regularization method and as a model sparsification technique. \citet{kingma_variational_2015}
consider Dropout \citep{hinton_improving_2012}, DropConnect \citep{wan_regularization_2013},
and Gaussian dropout \citep{srivastava_dropout_2014,wang_fast_2013} through the lens
of Bayesian inference methods and proposes \textit{Variational Dropout}. They argue that
the multiplicative noise introduced by these methods induces a distribution equivalent to
a fully factorized variational posterior of the form $
  q_\theta(\omega) = \prod_j q_{\theta}(\omega_j)
$, where $q_{\theta}(\omega_j)$ is $\omega_j = \mu_j \xi_j$ with $
  \xi_j \sim p_\theta(\xi_j)
$ iid from some $p_\theta(\xi)$.
%
% Binary dropout with rate $p \in (0, 1)$ uses $
%   p_\xi
%     = \mathcal{B}\bigl(
%       \{0, \tfrac1{1-p}\}, 1-p
%     \bigr)
% $, whereas in Gaussian dropout $
%   \xi \sim \mathcal{N}(1, \alpha)
% $ for $\alpha = \tfrac{p}{1-p}$.
% This suggests making $\alpha$ a variational parameter
% and optimizing it in \eqref{eq:elbo_general} with an appropriate penalty term. Thus,
%
Variational Dropout assumes fully factorized Gaussian approximate posterior $
  q_\theta(\omega)
    = \prod_j \mathcal{N}(\mu_j, \alpha_j \mu_j^2)
$ and factorized log-uniform prior $\pi(\omega)$ with $
  \pi(\omega_j) \propto \lvert \omega_j \rvert^{-1}
$. The divergence term in \eqref{eq:elbo} unravels into a sum of the following
analytically intractable terms:
\begin{equation}  \label{eq:improper-kl-div-real}
  KL(q_\theta \| \pi)
    \propto
      \frac12 \mathbb{E}_{\varepsilon \sim \mathcal{N}(0, 1)}
        \log{\Bigl\lvert \tfrac1{\sqrt{\alpha}} + \varepsilon \Bigr\rvert^2}
      % \frac12 \mathbb{E}_{\varepsilon \sim \mathcal{N}(0, 1)}
      %   \log{\bigl\lvert 1 + \sqrt{\alpha} \varepsilon \bigr\rvert^2}
      % - \frac12 \log{\alpha}
  \,.
\end{equation}
\citet{kingma_variational_2015} approximate this expression by a non-linear polynomial
regression for truncated $\alpha \in (0, 1)$. \citet{molchanov_variational_2017}
refine the approximation by using a nonlinear regression based on sigmoid and soft-plus.
% see p.6 sec 3.4 of kingma_variational_2015
We independently verify their approximation against Monte-Carlo estimates for $\alpha$
varying over a fine logarithmic grid and the exact expression for the derivative of
\eqref{eq:improper-kl-div-real}, which we derive in appendix~\ref{sub:real-chisq-grad}.
% (!) good for SGVB: gradient-based methods require unbiased gradient estimates
%
% (in) it is in poor taste to brag about this begin the first time anyone has derived a
% derivative of an intractable function.

@@

\begin{figure}[ht]
  \centering
  \includegraphics[draft,width=1.\linewidth]{../assets/kde-histograms-of-log-alpha-distribution.pdf}
  \caption{%
    Smoothed density of $\log\alpha$ for a convolutional and dense layer from the
    model in sec.~\ref{sub:mnist_like_datasets}.
  }
  \label{fig:log-alpha__mnist-like}
\end{figure}

% on chosing the threshold
The threshold $\tau$ is picked so, that the remaining \textit{``non-zero''} parameters
are within $\delta$ relative tolerance of their mean with high probability under the
approximate posterior $q_\theta(W)$. Since \eqref{eq:c-gauss-vi} is Gaussian, $
  \tfrac{\lvert w - \mu \rvert^2}
        {\alpha \lvert \mu \rvert^2}
$ is $\chi^2_k$ distributed with shape $k=1$ ($\real$) or $2$ ($\cplx$), whence
\begin{equation} \label{eq:prob_relevant}
  \Pr\bigl(
    \lvert w - \mu \rvert \leq \delta \lvert \mu \rvert
  \bigr)
    = \Pr\biggl(
      \frac{\lvert w - \mu \rvert^2}{\alpha \lvert \mu \rvert^2}
        \leq \frac{\delta^2}{\alpha}
    \biggr)
    = \chi^2_k\biggl(\frac{\delta^2}{\alpha}\biggr)
  \,.
\end{equation}
% Figure~\ref{fig:log-alpha__mnist-like} (bottom) shows the probability of a $\real$
% (or $\cplx$)  parameter being non-zero for various $\delta\in(0, \tfrac12)$ as a function
% of $\log \alpha$.
For a reasonably generous $\delta = 25\%$ tolerance all values of $\log \alpha$ below
$-2.5$ yield a relatively high chance of a non-zero parameter. Given that $q_\theta$
is merely an approximation, we opt for a somewhat higher $\log \alpha$ threshold, and
pick $\tau \in [-\tfrac12, \tfrac32]$, which is lower than $\tau = 3$, recommended by
\citet{molchanov_variational_2017}.

@@

% mention the same experimental setup as in MNIST
Experiments on CIFAR10 dataset are set up similarly to MNIST-like experiment, except
that we study VGG16 model \citep{simonyan_very_2015} and its $\cplx$-counterpart, fit
on the full training split of $50k$ colour $32\times32$ images, and use only raw image
input features (\texttt{raw}).

@@

\begin{figure}[!ht]
  \centering
  \begin{subfigure}[b]{0.9\textwidth}  % imcl2019-style swears at this
    \centering
    \includegraphics[draft,width=1.\linewidth]{../assets/grid_cplx_fine_kl_div_v3_fast-c-ard-trade-off.pdf}
    % \caption{} \label{fig:log-alpha-kde}
  \end{subfigure} \\
  % \hfill
  \begin{subfigure}[b]{0.9\textwidth}  % imcl2019-style swears at this
    \centering
    \includegraphics[draft,width=1.\linewidth]{../assets/grid_cplx_fine_kl_div_v3_fast-c-ard-trade-off.pdf}
    % \caption{} \label{fig:log-alpha-proba}
  \end{subfigure}
  \caption{%
    Pre ``fine-tune'' performance of the best models from figure \ref{fig:musicnet_trade-off}
    compressed with methods of sec.~\ref{sec:dropout_for_c_valued_layers}:
    (\ref{ssub:variational_dropout} -- \textit{top}, \ref{ssub:empirical_bayes} --  \textit{bottom}).
  }
  \label{fig:musicnet_trade-off_pre_fune-tune}
\end{figure}
% subsection muscinet (end)
